\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-09 chapters/abstract.tex
Self-plagiarism check

\paragraph{Keywords:} \acl{bci}, gaze independence, covert
visuospatial attention, event-related potentials, linear decoding, multilinear
decoding, motor impairment, gaze impairment

>>>>

\paragraph{Keywords:} brain-computer interface, electroencephalography,
event-related potentials, covert visuospatial attention, (multi)linear
decoding, severe physical impairment

### 2024-10-09 chapters/abstract_fr.tex
Self-plagiarism check

\paragraph{Mots-clés :} interface cerveau-ordinateur, indépendance du regard, attention visuo-spatiale cachée, potentiels évoqués liés aux événements, décodage linéaire, décodage multilinéaire, handicap moteur, handicap du regard.

>>>>

\paragraph{Mots-clés :} interface cerveau-ordinateur, électroencéphalographie,potentiels évoqués
attnetion visuo-spatiale cachée, decodage (multi-)linéaire, patients lourdements handicapées

### 2024-10-09 chapters/abstract_nl.tex
Self-plagiarism check

Personen met ernstige motorische beperkingen, zoals degenen met het
Locked-in Syndroom (LIS) of oog-motorische disfuncties, ondervinden
aanzienlijke uitdagingen bij het gebruik van traditionele visuele brain-computer interfaces (BCI's).
Deze systemen vereisen dat gebruikers hun blik richten op specifieke doelen, een taak die
onuitvoerbaar wordt voor personen met beperkte of
geen controle over oogbewegingen.
Dit werk richt zich op deze beperking door
BCI-methoden onafhankelijk van oogbewegingen te ontwikkelen, met een focus op
het verbeteren van de bedekte visuospatiële aandacht (VSA), waarbij gebruikers
hun aandacht kunnen richten op een doel zonder overeenkomstige oogbewegingen.
Een belangrijke bijdrage van dit werk is het compenseren van variabiliteit in
de timing van event-gerelateerde potentialen (ERP's), wat de decodeerprestaties
van bedekte VSA negatief beïnvloedt.
Door deze variabiliteit aan te pakken, verbetert de voorgestelde methode de
nauwkeurigheid van de communicatie zonder afhankelijkheid van oogcontrole,
waardoor BCI's bruikbaarder worden voor personen met motorische beperkingen van
de ogen.

@@

\paragraph{Trefwoorden:} brain-computer interface, blikonafhankelijkheid, verborgen visuospatiële aandacht, event-gerelateerde potentialen, lineaire decodering, multilineaire decodering, motorische beperking, blikstoornis.

>>>>

Personen met ernstige motorische beperkingen, zoals degenen met het
Locked-in Syndroom (LIS) of oog-motorische disfuncties, ondervinden
aanzienlijke uitdagingen bij het gebruik van traditionele visuele brein-computer interfaces (BCI's).
Deze systemen vereisen dat gebruikers hun blik richten op specifieke doelen, een taak die
onuitvoerbaar wordt voor personen met beperkte of
geen controle over oogbewegingen.
Dit werk richt zich op deze beperking door
BCI-methoden onafhankelijk van oogbewegingen te ontwikkelen, met een focus op
het verbeteren van de bedekte visuospatiële aandacht (VSA), waarbij gebruikers
hun aandacht kunnen richten op een doel zonder overeenkomstige oogbewegingen.
Een belangrijke bijdrage van dit werk is het compenseren van variabiliteit in
de timing van event-gerelateerde potentialen (ERP's), wat de decodeerprestaties
van bedekte VSA negatief beïnvloedt.
Door deze variabiliteit aan te pakken, verbetert de voorgestelde methode de
nauwkeurigheid van de communicatie zonder afhankelijkheid van oogcontrole,
waardoor BCI's bruikbaarder worden voor personen met motorische beperkingen van
de ogen.

@@

\paragraph{Trefwoorden:} brein-computer interface, electroencephalografie,
event-gerelateerde potentialen, bedekte visuospatiale attentie, (multi)lineaire
decodering, zware fysieke beperkingen

### 2024-10-09 chapters/bci.tex
Self-plagiarism check

\chapter{Brain-computer interfaces}
\label{sec:bci}
\todo{consistently terminology individuals with severe
speech and physical impairments (SSPI)}
\todo{uniform vsa condition vsa setting}
\todo{are all acronyms defined/Acro corectly used in chapters?}
\todo{footnot skips a page in wcble chap}

@@

\section{A direct interface to the brain}
\todohsm{make less verbose}

@@

The recorded brain activity forms only one part of the interaction scheme presented in
\cref{fig:bci/loop}.
It would make little sense to start digging around at random in all the brain activity
to try to extract the desired output.
Instead, it would help greatly if we knew which specific activity to look for.
A \ac{bci} operates according to a specified \emph{paradigm}, which defines the method
by which interaction is conducted.
This usually means that the user is instructed to perform a specific task, like
attending a specific flickering stimulus.
We can now exploit background knowledge from prior neuroscientific research to couple
specific activity in the brain to specific information or actions we want to convey via
the \ac{bci}.
The brain signal now \emph{modulates} this information.
In turn, the detection of this signal can be coupled to an action, like typing the
intended letter A or B in \cref{fig:bci/loop}.
The \ac{bci} paradigm comprises the manner of stimulation (if applicable) and the exact
task performed by the user, and is often closely linked to the manner of feedback in the
case of a closed-loop system.
As shown in \cref{sec:bci/paradigms}, there is a wide variety of \ac{bci} paradigms
based on different systems in the brain.

The previous points to another component of a successful \ac{bci}: the specific brain
signatures related to our paradigm need to be identified within the recorded activity.
The electrical activity of the brain is relatively weak compared to that of the
environment around us, containing electronic devices and other sources of interference.
Furthermore, the brain is continuously processing information and carrying out
`background tasks,' all of which generate their own brain activity.
On the other hand, the activity of interest often originates from a specific region or
network within the brain.
This activity might not easily be discerned from all other brain and environmental
activity.
Compare this to trying to listen to a single speaker at a noisy party where everyone is
shouting over each other.
Some obvious sources of interference can be easily filtered out by \emph{preprocessing}
(see \cref{sec:bci/preprocessing}) and by choosing the correct representation of the
signal (i.e., \emph{feature extraction}), but the problem of retrieving or
\emph{decoding} only the relevant signal is generally hard.
Nevertheless, we can attempt to solve it through supervised machine learning.
Useful information that is encoded in the brain signal by the paradigm can be retrieved
from it by the decoder and interpreted in function of the desired \ac{bci} output.
\Cref{sec:bci/decoding} takes a closer look at some common decoding techniques.

Finally, the loop can be closed by coupling this to a device or actuator.
There are two aspects to this.
On the one hand, the \ac{bci} gains its function by allowing the user to interact with
their environment.
For a communication system, this means coupling the decoded information to, e.g., a
virtual keyboard or speech synthesizer.
On the other hand, the actions performed by the \ac{bci} can themselves alter the user's
cognitive state or brain activity, creating an adaptive system.
A direct form of closed-loop \ac{bci} is the use of a neurostimulator.
Examples of this are deep brain stimulation to mitigate the symptoms of Parkinson's
disease, or the cochlear implant as a hearing prosthesis.
More indirectly, modulation can be achieved by sensory stimulation.
This involves specific, paradigm-related sensory input, like tactile feedback for
movement \ac{bci}~\cite{Flesher2021}, or simply presenting selected actions back to the
user.
The user's brain will then adapt through reinforcement learning, causing changes in
behavior or strategy.
Neuroplasticity, the brain's ability to adapt and reinforce specific neural pathways,
can have a positive impact on \ac{bci} performance and gives rise to rehabilitation
applications.

@@

The epochs contain values representing the voltage of a specific channel relative to the
reference at a specific time relative to the stimulus.
In \ac{erp} classification, they are usually directly used as features, by concatenating
all channels in the matrix format epoch $\mat{X}$ into one feature vector such that
$\mat{x} = \vec(\mat{X})$.

>>>>

\chapter{Brain-computer interfaces}
\label{sec:bci}

@@

\section{A direct interface to the brain}

@@

The recorded brain activity forms only one part of the interaction scheme presented in
\cref{fig:bci/loop}.
It would be inefficient to search all brain activity randomly for the desired output.
Instead, knowing the specific activity to target would help significantly.
A \ac{bci} operates under a specified \emph{paradigm}, which defines how interaction
is conducted.
This typically means instructing the user to perform a task, like attending to a
flickering stimulus.
Background knowledge from neuroscience research allows us to couple specific brain
activity with the information or actions conveyed via the \ac{bci}.
The brain signal now \emph{modulates} this information.
In turn, detecting this signal can trigger an action, like typing the intended letter A
or B in \cref{fig:bci/loop}.
The \ac{bci} paradigm includes both the method of stimulation (if applicable) and the
user's task, and is often closely tied to the feedback mechanism in a closed-loop system.
As shown in \cref{sec:bci/paradigms}, \ac{bci} paradigms vary widely based on different
brain systems.

Another critical component of a successful \ac{bci} is identifying brain signatures
related to the paradigm within the recorded activity.
The brain's electrical activity is weak compared to environmental noise from electronic
devices and interference.
Additionally, the brain continuously processes information and carries out
`background tasks', which generate more brain activity.
The desired activity often originates from specific brain regions or networks.
This signal can be difficult to isolate from all other brain and environmental activity.
It is similar to trying to hear a single speaker at a noisy party where everyone is
shouting.
Some interference can be filtered out through \emph{preprocessing}
(see \cref{sec:bci/preprocessing}) and by selecting the correct signal representation
(i.e., \emph{feature extraction}), but retrieving or \emph{decoding} the relevant signal
remains challenging.
Supervised machine learning can help solve this.
The information encoded in the brain signal can be retrieved by the decoder and used
to determine the desired \ac{bci} output.
\Cref{sec:bci/decoding} reviews some common decoding techniques.

Finally, the loop can be closed by coupling this to a device or actuator.
There are two aspects to this.
On the one hand, the \ac{bci} gains its function by allowing the user to interact with
their environment.
For a communication system, this means coupling the decoded information to, e.g., a
virtual keyboard or speech synthesizer.
On the other hand, the actions performed by the \ac{bci} can themselves alter the user's
cognitive state or brain activity, creating an adaptive system.
This happens either directly through electrical neurostimulation, as is the
case for adaptive deep brain stimulation in Parkinson's disease or for the
cochlear implant as a hearing aide, or indirectly through sensory stimulation.
The latter involves specific, paradigm-related, sensory input, like tactile feedback for
movement \ac{bci}~\cite{Flesher2021}, or simply presenting selected actions back to the
user.
The user's brain will then adapt through reinforcement learning, causing changes in
behavior or strategy.
Neuroplasticity, the brain's ability to adapt and reinforce specific neural pathways,
can have a positive impact on \ac{bci} performance and gives rise to rehabilitation
applications.

@@

The epochs contain values representing the voltage of a specific channel relative to the
reference at a specific time relative to the stimulus.
In \ac{erp} classification, they are usually directly used as features, by concatenating
all channels in the matrix format epoch $\mat{X}$ into one feature vector such that
$\mat{x} = \text{vec}(\mat{X})$.
